---
title: "Lecture 6.1 TEMPLATE"
subtitle: "Typography"
author: "Your Name"
date: 2025-02-10
toc: true
---

::: callout-note
This template follows [lecture 6.1 slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture6.1-typography-slides.html#/title-slide){target="_blank"}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

## System fonts

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
# import packages ---- 
library(palmerpenguins)
library(tidyverse)

# create ggplot to demonstrate use of system fonts ----
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point() + 
  labs(title = "This title is serif font",
       subtitle = "This subtitle is mono font",
       x = "This axis label is sans font\n(default)",
       y = "This axis is also sans font\n(default)") +
  theme(
    plot.title = element_text(family = "serif", size = 30),
    plot.subtitle = element_text(family = "mono", size = 25),
    axis.title = element_text(family = "sans", size = 22),
    axis.text.x = element_text(family = "serif", face = "bold", size = 18),
    axis.text.y = element_text(family = "mono", face = "italic", size = 18)
    )
```

## Setup

::: callout-note
## Recall our original plot from [lecture 4.1](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture4.1-amounts-slides.html#/title-slide){target="_blank"}

Our "first draft" of this plot began in week 4 during our amounts / rankings lecture. We'll focus on improving colors and fonts today.
:::

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(showtext)
library(glue)
library(ggtext)

#......................import Google fonts.......................
# `name` is the name of the font as it appears in Google Fonts
# `family` is the user-specified id that you'll use to apply a font in your ggpplot

font_add_google(name = "Josefin Sans", 
                family = "josefin") #loading from google fonts, fam = what you want to call it 
font_add_google(name = "Sen",
                family = "sen")


#....................import Font Awesome fonts...................

font_add(family = "fa-brands", #calling from downloaded font file, calling it fa-brands 
         regular = here::here("fonts", "Font Awesome 6 Brands-Regular-400.otf")) 

font_add(family = "fa-regular", #calling from downloaded font file 
         regular = here::here("fonts", "Font Awesome 6 Free-Regular-400.otf")) 

font_add(family = "fa-solid", #calling from downloaded font file 
         regular = here::here("fonts", "Font Awesome 6 Free-Solid-900.otf")) 

#......enable {showtext} rendering for all newly opened GDs......

showtext.auto() #turn on show text, need to do before loading text 
showtext_opts(dpi = 300)

#..........................import data...........................
# find import code at: https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-03-05#grab-the-clean-data-here
jobs <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/jobs_gender.csv")

```

## Wrangle data

See [lecture 4.1 slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture4.1-amounts-slides.html#/title-slide){target="_blank"} (specifically, [this slide](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture4.1-amounts-slides.html#/data-wrangling){target="_blank"} and [this slide](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture4.1-amounts-slides.html#/dumbbell-plot-subset){target="_blank"}) as a reminder of where we left off.

```{r}
#| eval: true
#| echo: true
#| warning: false
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                wrangle data                              ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

jobs_clean <- jobs |> 
  
  # add col with % men in a given occupation (% females in a given occupation is already included) ----
  mutate(percent_male = 100 - percent_female) |>
         
  # rearrange columns ----
  relocate(year, major_category, minor_category, occupation,
          total_workers, workers_male, workers_female,
          percent_male, percent_female,
          total_earnings, total_earnings_male, total_earnings_female, 
          wage_percent_of_male) |> 
  
  # drop rows with missing earnings data ----
  drop_na(total_earnings_male, total_earnings_female) |> 
  
  # make occupation a factor (for reordering groups in our plot) ----
  mutate(occupation = as.factor(occupation)) |> 
 
  # classify jobs by percentage male or female (these will become facet labels in our dumbbell plot) ----
  mutate(group_label = case_when(
    percent_female >= 75 ~ "Occupations that are 75%+ female",
    percent_female >= 45 & percent_female <= 55 ~ "Occupations that are 45-55% female",
    percent_male >= 75 ~ "Occupations that are 75%+ male"
  )) 

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                              create subset df                            ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#....guarantee the same random samples each time we run code.....
set.seed(0)

#.........get 10 random jobs that are 75%+ female (2016).........
f75 <- jobs_clean |> 
  filter(year == 2016, group_label == "Occupations that are 75%+ female") |> 
  slice_sample(n = 10)

#..........get 10 random jobs that are 75%+ male (2016)..........
m75 <- jobs_clean |> 
  filter(year == 2016, group_label == "Occupations that are 75%+ male") |> 
  slice_sample(n = 10)

#........get 10 random jobs that are 45-55%+ female (2016).......
f50 <- jobs_clean |> 
  filter(year == 2016, group_label == "Occupations that are 45-55% female") |> 
  slice_sample(n = 10)

#.......combine dfs & relevel factors (for plotting order).......
subset_jobs <- rbind(f75, m75, f50) |> 
  mutate(group_label = fct_relevel(.f = group_label, 
                                   "Occupations that are 75%+ female", 
                                   "Occupations that are 45-55% female", 
                                   "Occupations that are 75%+ male"),
         occupation = fct_reorder(.f = occupation, .x = total_earnings))
```

## Recreate original plot

See [lecture 4.1](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture4.1-amounts-slides.html#/dumbbell-plot-create){target="_blank"} as a reminder.

```{r}
#| eval: true
#| echo: true
#| warning: false
#| out-width: "100%"
# recreate original plot ----
plot <- ggplot(subset_jobs) +
  
  # create dumbbells ----
  geom_linerange(aes(y = occupation,
                     xmin = total_earnings_female, xmax = total_earnings_male)) + 
  geom_point(aes(x = total_earnings_male, y = occupation), 
             color = "#CD93D8", 
             size = 2.5) +
  geom_point(aes(x = total_earnings_female, y = occupation), 
             color = "#6A1E99", 
             size = 2.5) +
  
  # facet wrap by group ----
  facet_wrap(~group_label, nrow = 3, scales = "free_y") + # "free_y" plots only the axis labels that exist in each group
  
  # axis breaks & $ labels ----
  scale_x_continuous(labels = scales::label_currency(scale = 0.001, suffix = "k"),
                     breaks = c(25000, 50000, 75000, 100000))

plot
```

## Create new palette

```{r}
#| eval: true
#| echo: true
#| warning: false
# create a named color palette ----
# to avoid some annoying typing, feel free to copy these into our palette vector:
# "males" = "#2D7787", "females" = "#FC6B4B",  "dark_text" = "#0C1509", "light_text" = "#4E514D"

earnings_pal <- c("males" = "#2D7787", 
                  "females" = "#FC6B4B",  
                  "dark_text" = "#0C1509", #used to color text elements later 
                  "light_text" = "#4E514D")

# preview it ----

monochromeR::view_palette(earnings_pal) #prints out palette with variable name too, super nice
```

## Update plot colors & fonts

```{r}
#| eval: true
#| echo: true
#| warning: false
#| out-width: "100%"
#.........................create caption.........................

github_icon <- "&#xf09b" #&#x are to tell r to look for unicode which we got from font awesome
github_username <- "slance99"

caption <- glue::glue(
  "Data Source: TidyTuesday (March 5, 2019) <br> 
  <span style ='font-family:fa-brands;'>{github_icon}; </span>
  {github_username}" #call two icons above 
)

#........................create subtitle.........................
money_icon <- "&#xf3d1"

subtitle <- glue::glue("<span style='font-family:fa-regular;'>{money_icon};</span>
                       Median earnings of full-time 
                       <span style='color: #2D7787;'> **male** </span>
                       male 
                       versus 
                       <span style='color: #FC6B4B;'> **female** </span>
                       female workers by occupation in 2016")

#..........................updated plot..........................
plot <- ggplot(subset_jobs) +
  
  # create dumbbells ----
  geom_linerange(aes(y = occupation,
                     xmin = total_earnings_female, xmax = total_earnings_male)) + 
  geom_point(aes(x = total_earnings_male, y = occupation), 
             color = earnings_pal["males"], #updating colors to index by males 
             size = 2.5) +
  geom_point(aes(x = total_earnings_female, y = occupation), 
             color = earnings_pal["females"], 
             size = 2.5) +
  
  # facet wrap by group ----
  facet_wrap(~group_label, nrow = 3, scales = "free_y") + # "free_y" plots only the axis labels that exist in each group
  
  # axis breaks & $ labels ----
  scale_x_continuous(labels = scales::label_currency(scale = 0.001, suffix = "k"),
                     breaks = c(25000, 50000, 75000, 100000)) +
  labs(title = "Males earn more than females across occupations",
       subtitle = subtitle, 
       caption = caption) +
  theme_minimal(base_size = 5) + #set a base size and set relative sizes of all font 
  theme(
    plot.title.position = "plot", #moves everything over to the left hand side 
    plot.title = element_text(family = "josefin", #font family imported from google 
                              face = "bold",
                              size = rel(1.7),
                              color = earnings_pal["dark_text"]),
    plot.subtitle = ggtext::element_textbox(family = "sen",
                                 size = rel(1.1),
                                 color = earnings_pal["light_text"],
                                 width = unit(30, "cm"), #text too wide for dpi setting 
                                 margin = margin(t = 2, r = 0, b = 6, l = 0 ),
                                 padding = margin(t=5, r=0, b=5, l=0)), #plot breathinh
    plot.caption = ggtext::element_textbox(family = "sen",
                                face = "italic", #sen does not take italic 
                                color = earnings_pal["light_text"],
                                margin = margin(t = 15),
                                width = unit(25, "cm"), 
                                halign = 1,
                                lineheight = 1.5,
                                padding = margin(t=10, r=0, b=10, l=0)),
    strip.text = element_text(family = "josefin",
                              face = "bold", #text above each plot 
                              size = rel(1),
                              hjust = 0),
    panel.spacing.y = unit(0.5, "cm"), #space out the panels 
    axis.text = element_text(family = "sen",
                             size = rel(1), #change what the axis text looks like 
                             color = earnings_pal["light_text"]),
    plot.background = element_rect(color = "white"), #make background white 
    axis.title = element_blank() #get rid of individual plot titles 
  )

plot
```

## Save plot as a PNG file

```{r}
#| eval: true
#| echo: true
# write plot to file (aka save as png) ----

ggsave(
  filename = here::here("week_6", "images", "salary.png"),
  plot = plot,
  device = "png",
  width = 5, 
  height = 5,
  unit = "in",
  dpi = 300
)
```

## Turn of `{showtext}` text rendering

```{r}
# turn off {showtext} text rendering ----
showtext_auto(FALSE)
```
