---
title: "Lecture 4.2 TEMPLATE"
subtitle: "Visualizing numerical relationships"
author: "Your Name"
date: 2025-01-27
toc: true
editor_options: 
  chunk_output_type: inline
---

::: callout-note
This template follows [lecture 4.2 slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture4.2-relationships-slides.html#/title-slide){target="_blank"}.
:::

## Setup

```{r}
#| eval: false
#| echo: true
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(metajam) 
library(tidyverse)
library(ggExtra)
library(ggdensity)
library(here)


#........................get the data file.......................
stream_chem_data <- read_csv(here("week_4/data/HubbardBrook")))
```

## Scatter plots

### Basic scatter plot

```{r}
#| eval: true
#| echo: true
#| warning: false
# basic scatter plot ----
```

### Alt 1: add a rug plot

```{r}
#| eval: true
#| echo: true
#| warning: false
# scatter + rug plot ----
p1
```

### Alt 2: marginal plot alternatives using `{ggExtra}`

```{r}
#| eval: true
#| echo: true
#| warning: false
# add marginal plots with {ggExtra}; also try "density", "boxplot" ----
```

### Alt 3: scatter + marginal plot with 2+ groups

```{r}
#| eval: true
#| echo: true
#| warning: false
# scatter plot with points colored by site ----
p2 <- stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH)) + 
  geom_point(alpha = 0.5)

p2
# add marginal density plot ----
```

## Trend lines

### Default (\<1000 data points): LOESS method

-   similar to a moving average

```{r}
#| eval: true
#| echo: true
# add a smoothed line ----
```

### Line of best fit

```{r}
#| eval: true
#| echo: true
# add a trend line ----
stream_chem_data |> 
  filter(waterYr %in% c(2011:2021)) |>  
  filter(site == "W5") |> 
  ggplot(aes(x = DOC, y = pH)) + 
  geom_point(alpha = 0.5) +
  geom_smooth()
```

## Visualizing a third numeric variable

### Bubble charts

-   variation of a scatter plot for visualizing a third numeric variable
-   be mindful: it's challenging to interpret 3 numeric variables

```{r}
#| eval: true
#| echo: true
#| warning: false
# bubble chart ----
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH)) + 
  geom_point(alpha = 0.5) +
  labs(x = "DOC (mg/L)", size = "Al (mg/L)", color = "Site")
```

-   adjust size range of bubbles

```{r}
#| eval: true
#| echo: true
#| warning: false
# scale size by area ----
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH, color = site, size = Al_ICP)) + 
  geom_point(alpha = 0.5) +
  labs(x = "DOC (mg/L)", size = "Al (mg/L)", color = "Site")
```

-   IMPORTANT: don't scale size by radius (can be deceptive)

```{r}
#| eval: true
#| echo: true
#| warning: false
# don't scale size by radius ----
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH, color = site, size = Al_ICP)) + 
  geom_point(alpha = 0.5) +
  scale_radius(range = c(1, 10)) +
  labs(x = "DOC (mg/L)", size = "Al (mg/L)", color = "Site")
```

### Use color for third numeric variable

```{r}
#| eval: true
#| echo: true
#| warning: false
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH)) + 
  geom_point(alpha = 0.5, size = 2) +
  labs(x = "DOC (mg/L)")
```

### But oftentimes best to just create two different plots

```{r}
#| eval: true
#| echo: true
#| warning: false
#| layout-ncol: 2
# effect of DOC on pH ----
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = DOC, y = pH, color = site)) + 
  geom_point(alpha = 0.5)

# effect of DOC on Al_ICP ----
stream_chem_data |> 
  filter(waterYr == 2021) |> 
  ggplot(aes(x = Al_ICP, y = pH, color = site)) + 
  geom_point(alpha = 0.5)
```

## Overplotting

-   if you have too many points, scatter plots can be ineffective

```{r}
#| eval: true
#| echo: false
#| warning: false
# wayyyy too many points ----
```

### Initial strategies

-   adjust size / transparency of points
-   add a rug plot (or marginal plot)
-   color by group

```{r}
#| eval: true
#| echo: false
#| warning: false
# smaller + more transparent points ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) + 
  geom_point() 
```

### Alt 1: 2d heatmaps

-   e.g. like you're looking down on a histogram

```{r}
#| eval: true
#| echo: false
#| warning: false
# heatmap of 2d bin counts ----
ggplot(stream_chem_data, aes(x = SO4, y = pH))
```

-   or use hexagonal shapes
-   also increase height of legend can make continuous scale easier to read

```{r}
#| eval: true
#| echo: false
#| warning: false
# heatmap of 2d bin counts ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) 
```

### Alt 2: 2d density / contour plots

-   e.g. like you're looking down on a density plot
-   legend provides an estimate of the proportion of data points fall within a colored region (density of distribution of points sums to 1)
    -   interpreting legend: 0-2% of of points fall within a 1x1 square in the darkest blue region, while 26-28% fall within a 1x1 square in the brightest yellow region

```{r}
#| eval: true
#| echo: false
#| warning: false
# contour plot ----
ggplot(stream_chem_data, aes(x = SO4, y = pH))
```

### Alt 3: 2d density plots using `{ggdensity}`

-   can be easier to interpret
-   compute and plot the resulting highest density regions (HDRs)
-   HDRs are computed to be the smallest such regions that bound that level of probability

```{r}
#| eval: true
#| echo: false
#| warning: false
#  ----
ggplot(stream_chem_data, aes(x = SO4, y = pH)) 
```
